The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-05-07 paper.qmd
edited background

Uncertainty visualisation is a relatively new field in research. Early papers that specifically reference "uncertainty visualisation" appear in the late 80s [@Ibrekk1987], with geospatial information visualisation literature in the early 90s declaring this to be essential aspect of information display [@MacEachren1992; @Carr1992]. Despite the new terminology visualisation of uncertainty has been present since the earliest times. For example, box plots or histograms can be considered to be displaying uncertainty in the sense of variability in observations sampled from a population distribution.

@@

# Display Plots
p1
p2
p3
p4
p5
p6
p7
p8
p9
```
However, uncertainty visualisation takes many forms today, It can be considered to expand the application of data visualisation towards making better decisions, and research is concerned about how the construction of plots and information contained might affect the resulting decisions. With the abundance of publications it is timely to consider a review of the state-of-the-art. In fact, there have already been several reviews published.

@@

The review provided here attempts to address these issues. We begin with understanding the definition problem, to obtain and understanding of the *purpose* of visualising uncertainty, and understand if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for new methodology. We also posit that "uncertainty visualisation" is too diffuse to be considered a field in itself, but rather should be dissected into smaller elements. The ideal role of this review to help synthesize this noisy field, by collecting together essential components, summarising different viewpoints the current literature, and provide a guide for potential research topics.

>>>>

Uncertainty visualisation is a relatively new field in research. Early papers that specifically reference "uncertainty visualisation" appear in the late 80s [@Ibrekk1987], with geospatial information visualisation literature in the early 90s declaring this to be essential aspect of information display [@MacEachren1992; @Carr1992]. These early experiments typically involved showing participants a distribution, such as those depicted in @fig-ibrekk, and asking the viewers to extract a probability or average. Despite the new terminology visualisation of uncertainty has been present since the earliest times. For example, box plots or histograms can be considered to be displaying uncertainty in the sense of variability in observations sampled from a population distribution.

@@

# Display Plots
p1
p2
p3
p4
p5
p6
p7
p8
p9
```

However, uncertainty visualisation takes many forms today, It can be considered to expand the application of data visualisation towards making better decisions, and research is concerned about how the construction of plots and information contained might affect the resulting decisions. With the abundance of publications it is timely to consider a review of the state-of-the-art. In fact, there have already been several reviews published.

@@

The review provided here attempts to address these issues. We highlight the three key issues facing the uncertainty visualisation literature, specifically the lack of a definition of uncertainty, the confusion as to what *is* an "uncertainty visualisation", and the difficulties with measuring "uncertainty". When disucssing each of these issues, we will discuss how this concept is understood in existing literature, why this understanding may be flawed, and the issues that arrise due to these flaws. By considering the *purpose* of visualizing uncertainty we avoid providing a simple list of the current methods, and instead consider if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for new methodology. The ideal role of this review to help synthesize this noisy field, by collecting together essential components, summarizing different viewpoints the current literature, and provide a guide for potential research topics.
